---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TOut

<!-- badges: start -->
<!-- badges: end -->

The goal of TOut is to optimise the design of **T**hree-**Out**come clinical trials by determining their sample size and their decision rule thresholds (also known as progression criteria in the field of pilot trials).

## Installation

You can install the development version of TOut from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("DTWilson/TOut")
```

## Binary outcome

### Fixed sample size

Suppose we want to use a pilot trial with $n$ participants per arm to decide if a larger, definitive trial will be feasible with respect to the level of adherence to treatment in the experimental arm. Adherence is a binary event for each participant, which will occur with probability $\rho$. We would like to avoid conducting the main trial if the adherence rate is less that 0.5, but would like to proceed if it is greater than 0.7. When adherence is between these two levels, we would ideally like to pause amd make our progression decision based on other observations made during the pilot trial. Our aim is to find optimal progression criteria $x_0, x_1$ such that

$$
Decision =
\begin{cases}
 stop    & \hat{\rho} \leq x_0 \\
 pause   & x_0 < \hat{\rho} \leq x_1 \\
 go      & x_1 < \hat{\rho}
 \end{cases}       
$$

To formalise this problem we set up null and alternative hypotheses as $\rho_0 = 0.5, \rho_1 = 0.7$ respectively, and define some operating characteristics which we want to minimise. Specifically, we will use:

- $\alpha$: the probability of proceeding to the main trial when $\rho = \rho_0$;
- $\beta$: the probability of not proceeding to the main trial when $\rho = \rho_1$; and
- $\gamma$: the probability of not obtaining a $pause$ decision when $\rho = (\rho_0 + \rho_1)/2$.

The TOut package allows us to specify maximum levels of these three operating characteristics, and then finds a design which will satisfy these constraints. When there are several such designs, it will suggest the one which minimises the average of the three operating characteristics.

```{r}
library(TOut)

opt_pc_bin(n = 100, rho_0 = 0.5, rho_1 = 0.7,
                     alpha_nom = 0.05, beta_nom = 0.2, gamma_nom = 0.5)
```

The function `opt_pc` returns a vector of the form $(n, x_0, x_1, \alpha, \beta, \gamma)$. In this case we find that a pilot trial with $n = 100$ patients in the intervention arm will satisfy our constraints $\alpha < 0.05, \beta < 0.2, \gamma < 0.5$ if we $stop$ when we observe 56 patients or fewer adhering, $go$ if we observe more that 68 patients adhering, and $pause$ otherwise.

### Optimising sample size

TOut can also find the lowest sample size which can satisfy all operating characteristic constraints:

```{r}
TOut_design_bin(rho_0 = 0.5, rho_1 = 0.7, alpha_nom = 0.05, beta_nom = 0.1, gamma_nom = 0.5)
```

Here we find that our earlier choice of $n = 100$ was unnecessarily large and can be reduced down to $n = 83$, with progression criteria $x_0 = 47$ and $x_1 = 54$.

## Continuous outcome

We can apply the same approach to the case of a continuous outcome. For example, suppose that our hypotheses are $\rho_0 = 0, \rho_1 = 0.3$ and our outcome has a standard deviation of $\sigma = 1$. The optimal sample size and progression criteria are then:

```{r}
TOut_design_cont(rho_0 = 0, rho_1 = 0.3, sigma = 1, alpha_nom = 0.05, beta_nom = 0.1, gamma_nom = 0.5)
```
In the continuous case, the progression criteria $x_0$ and $x_1$ are given on the scale of the z-statistic. 




