---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TOut

<!-- badges: start -->
<!-- badges: end -->

The goal of TOut is to optimise the design of Three-Outcome clinical trials by determining their sample size and their decision rule thresholds (also known as _progression criteria_ in the field of pilot trials).

## Installation

You can install the development version of TOut from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("DTWilson/TOut")
```

## Example

Suppose we want to use a pilot trial with $n$ participants per arm to decide if a larger, definitive trial will be feasible with respect to the level of adherence to treatment in the experimental arm. Adherence is considered to be a binary event for each participant, which will occur with probability $\rho$. We would like to avoid conducting the main trial if the adherence rate is less that 0.5, but would like to proceed when it is greater than 0.7. When adherence is between these two levels, we would like to make our progression decision based on other observations made during the pilot trial. Our aim is to find optimal progression criteria $x_0, x_1$ such that

$$
Decision =
\begin{cases}
 stop    & \hat{\rho} \leq x_0 \\
 pause   & x_0 < \hat{\rho} \leq x_1 \\
 go      & x_1 < \hat{\rho}
 \end{cases}       
$$

To formalise this problem we define a null and alternative hypothesis as $\rho_0 = 0.5, \rho_1 = 0.7$ respectively, and set up some operating characteristics which we want to minimise. Specifically, we will use:

- $\alpha$: the probability of proceeding to the main trial when $\rho = \rho_0$;
- $\beta$: the probability of not proceeding to the main trial when $\rho = \rho_1$; and
- $\gamma$: the probability of *not* obtaining a $pause$ decision when $\rho = (\rho_0 + \rho_1)/2$.

The TOut package allows us to specify maximum levels of these three operating characteristics, and then finds all designs (i.e. all possible choices of $x_0, x_1$) which will satisfy these constraints. When there are several such designs, it will suggest the one which minimises the average of the three operating characteristics.

```{r example}
library(TOut)

opt_pc(n = 200, rho_0 = 0.5, rho_1 = 0.7,
                     alpha_nom = 0.05, beta_nom = 0.2, gamma_nom = 0.5)
```
This function `opt_pc` returns a vector of the form $(n, x_0, x_1)$. In this case we find that a pilot trial with $n = 200$ patients in the intervention arm will satisfy our constraints $\alpha < 0.05, \beta < 0.2, \gamma < 0.5$ if we $stop$ if observing 109 patients or fewer adhere, $go$ if we observe more that 131 patients adhere, and $pause$ otherwise.
